---
title: "Education and Income Across American Cities"
author: "Amelia Alvarez"
description: "Income vs Education"
output:
  distill::distill_article:
    self_contained: false
---


## Introduction

High school graduation rates are one of the most important indicators of long term economic mobility and community well being. Cities with lower graduation rates tend to face higher unemployment and reduced civic participation. Understanding what drives differences in graduation rates across cities is therefore an important policy question.
This project examines how median household income at the city level is related to high school graduation rates across cities in Wisconsin and four other states chosen at random — New York, Colorado, Louisiana, and Wyoming.

My hypothesis is that cities with higher median household incomes will have higher high school graduation rates. This hypothesis is set in both behavioral and institutional explanations. Institutionally, wealthier communities tend to have better funded schools due to higher property taxes, resulting in smaller class sizes, more experienced teachers, and greater access to academic support. Behaviorally, economic hardship increases stress at home, may require students to work part time jobs, and reduces parental availability for academic support, all of which increase dropout risk. 

The explanatory variable being used is median household income, measured in dollars at the city level using the U.S. Census Bureau's American Community Survey 5-Year Estimates. This variable shows the variation across cities, ranging from low income urban areas to wealthy suburbs, which allows us to observe and study differences in income levels. The outcome variable is the high school graduation rate, measured as the percentage of adults 25 and older who have earned a high school diploma or equivalent in each city, also drawn from the American Community Survey. 

My hypothesis would be supported if the data show a positive relationship between median household income and graduation rates, meaning cities with higher incomes consistently show higher graduation rates. My hypothesis would be weakened if the data show no statistically significant relationship between the two variables, or if income loses its significance after controlling for other factors such as geographic location or city size.


```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(dplyr)
library(readr)
library(readxl)
library(ggplot2)
library(tidycensus)
```





```{r}
education <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP02_0068P",
  year = 2023,
  survey = "acs5"
)
```


```{r}
income <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
)
```


```{r api-key, echo = FALSE, message = FALSE, warning = FALSE}
census_api_key("1956434ce467efbcc6650c2a36eb9a5c4f758f22")
```


```{r, echo = FALSE, message = FALSE, warning= FALSE}

education <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(education = estimate)


income <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(income = estimate)

city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(NAME = gsub(", Wisconsin", "", NAME)) %>%
  filter(!is.na(education) & !is.na(income))

```


```{r plot-wisconsin, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8}
plot_wi <-ggplot(city_df, aes(x = education, y = income)) + 
  geom_point(color = "purple", size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income Across Wisconsin Cities",
    x = "% of Adults with a High School Diploma or Higher",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )

plot_wi
```

```{r, echo = FALSE, message = FALSE, warning= FALSE}

education <- get_acs(
  geography = "place",
  state = "NY",
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(education = estimate)


income <- get_acs(
  geography = "place",
  state = "NY",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(income = estimate)

city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(NAME = gsub(", New York", "", NAME)) %>%
  filter(!is.na(education) & !is.na(income))

```

```{r plot-new york, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8}
plot_ny <-ggplot(city_df, aes(x = education, y = income)) +
  geom_point(color = "blue", size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income Across New York Cities",
    x = "% of Adults with a High School Diploma or Higher",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )

plot_ny 
```



```{r, echo = FALSE, message = FALSE, warning= FALSE}

education <- get_acs(
  geography = "place",
  state = "WY",
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(education = estimate)


income <- get_acs(
  geography = "place",
  state = "WY",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(income = estimate)

city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(NAME = gsub(", Wyoming", "", NAME)) %>%
  filter(!is.na(education) & !is.na(income))

```

```{r plot-wyoming, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8}
plot_wy <-ggplot(city_df, aes(x = education, y = income)) +
  geom_point(color = "darkgreen", size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income Across Wyoming Cities",
    x = "% of Adults with a High School Diploma or Higher",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )

plot_wy
```






```{r, echo = FALSE, message = FALSE, warning= FALSE}

education <- get_acs(
  geography = "place",
  state = "CO",
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(education = estimate)


income <- get_acs(
  geography = "place",
  state = "CO",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(income = estimate)

city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(NAME = gsub(", Colorado", "", NAME)) %>%
  filter(!is.na(education) & !is.na(income))

```

```{r plot-colorado, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8}
plot_co <-ggplot(city_df, aes(x = education, y = income)) +
  geom_point(color = "darkorange", size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income Across Colorado Cities",
    x = "% of Adults with a High School Diploma or Higher",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )

plot_co
```




```{r, echo = FALSE, message = FALSE, warning= FALSE}

education <- get_acs(
  geography = "place",
  state = "LA",
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(education = estimate)


income <- get_acs(
  geography = "place",
  state = "LA",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(income = estimate)

city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(NAME = gsub(", Louisiana", "", NAME)) %>%
  filter(!is.na(education) & !is.na(income))

```

```{r plot-louisiana, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8}
plot_la <-ggplot(city_df, aes(x = education, y = income)) +
  geom_point(color = "cyan3", size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income Across Louisiana Cities",
    x = "% of Adults with a High School Diploma or Higher",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )
plot_la
```

```{r}

library(patchwork)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 19, fig.height = 12}
(plot_wi | plot_ny | plot_wy) /
(plot_co | plot_la)
```


## Analysis

The scatter plots above display the relationship between high school graduation rates and median household income across the cities. Across all five states, a clear positive trend is visible. The data shows cities with higher percentages of adults holding a high school diploma tend to have higher median household incomes. This pattern is consistent with my hypothesis and is supported by the upward sloping trend line in every state.

With all these state data taken together, the evidence consistently supports my hypothesis that cities with higher median household incomes tend to have higher high school graduation rates. The relationship appears in every state regardless of geographic region, state size, or overall income level, suggesting it is a robust and systematic pattern rather than a coincidence specific to one location.



```{r load-libraries, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(modelsummary)
```

```{r load-data, echo = FALSE, message = FALSE, warning = FALSE}

states <- c("WI", "NY", "WY", "CO", "LA")


education <- get_acs(
  geography = "place",
  state = states,
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(hs_graduation_rate = estimate)


income <- get_acs(
  geography = "place",
  state = states,
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(median_income = estimate)


city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(
    # Extract state name from the NAME column
    state = case_when(
      grepl("Wisconsin", NAME)   ~ "Wisconsin",
      grepl("New York", NAME)    ~ "New York",
      grepl("Wyoming", NAME)     ~ "Wyoming",
      grepl("Colorado", NAME)    ~ "Colorado",
      grepl("Louisiana", NAME)   ~ "Louisiana"
    ),
    
    NAME = gsub(", Wisconsin|, New York|, Wyoming|, Colorado|, Louisiana", "", NAME)
  ) %>%
  filter(!is.na(hs_graduation_rate) & !is.na(median_income) & !is.na(state))
```

```{r}

library(kableExtra)
```


```{r summary-table, echo = FALSE, message = FALSE, warning = FALSE}

city_df %>%
  group_by(state) %>%
  summarise(
    Cities = n(),
    `Avg HS Graduation Rate` = round(mean(hs_graduation_rate), 1),
    `Avg Median Income` = round(mean(median_income), 0)
  ) %>%
  kable(
    caption = "Summary of Data by State",
    col.names = c("State", "Number of Cities", "Average HS Graduation Rate (%)", "Average Median Household Income ($)"),
    format.args = list(big.mark = ",")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "bordered"),
    full_width = TRUE
  )
```







```{r regression, echo = FALSE, message = FALSE, warning = FALSE}

model <- lm(median_income ~ hs_graduation_rate, data = city_df)


modelsummary(
  model,
  stars = TRUE,
  coef_rename = c(
    "hs_graduation_rate" = "High School Graduation Rate (%)",
    "(Intercept)" = "Intercept"
  ),
  gof_map = c("nobs", "r.squared"),
  fmt = 0,
  title = "Regression: Effect of High School Graduation Rate on Median Household Income"
)
```