---
title: "PS270 Final Project"
author: "Amelia Alvarez"
description: "Income vs Education"
output:
  distill::distill_article:
    self_contained: false
---


## Project Thoughts

I am interested in exploring data related to high school graduation rates in areas of poverty. I would like to look at low income areas and see if they have lower graduation rates compared to more affluent areas.

## Project Hypothesis

My research question examines how city income level is related to high school graduation rates. Are students in low-income areas less likely to graduate from high school compared to students in higher-income areas? My hypothesis is that high school students in low-income areas are less likely to graduate from high school than students in higher-income areas. This hypothesis is grounded in the thought that schools in low-income areas often receive fewer financial resources due to property tax-based funding structures. This can result in larger class sizes, fewer experienced teachers, limited access to advanced coursework, fewer counselors, and academic support services. From a behavioral perspective, economic hardship can also increase stress at home, require students to work part-time jobs, or reduce parental availability for academic support. These disadvantages systematically reduce academic performance and increase dropout risk. My explanatory variable of interest is the socioeconomic status of the area. This can be measured by median household income in the school district or ZIP code, the percentage of students eligible for free or reduced-price lunch, and the poverty rate in the census tract or district. These measures vary across different areas, which allows us to observe variation across schools or districts in income levels. The outcome variable of interest is high school graduation rates. This can be measured by the four-year cohort graduation rate, which is the percentage of students who graduate within four years of entering ninth grade, or by looking at the dropout rate. My hypothesis would be supported if the data show a negative relationship between area income and graduation rates and if schools in lower-income areas consistently show lower graduation rates. My hypothesis would be weakened if the data show no relationship between area income and graduation rates, or if, after controlling for other variables such as school size, urban or rural location, and racial composition, income is no longer statistically significant, which would challenge the source of my hypothesis.


```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(dplyr)
library(readr)
library(readxl)
library(ggplot2)
library(tidycensus)
```





```{r}
education <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP02_0068P",
  year = 2023,
  survey = "acs5"
)
```


```{r}
income <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
)
```


```{r api-key, echo = FALSE, message = FALSE, warning = FALSE}
census_api_key("1956434ce467efbcc6650c2a36eb9a5c4f758f22")
```


```{r, echo = FALSE, message = FALSE, warning= FALSE}

education <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(education = estimate)


income <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(income = estimate)

city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(NAME = gsub(", Wisconsin", "", NAME)) %>%
  filter(!is.na(education) & !is.na(income))

```


```{r plot-wisconsin, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8}
plot_wi <-ggplot(city_df, aes(x = education, y = income)) + 
  geom_point(color = "purple", size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income Across Wisconsin Cities",
    x = "% of Adults with a High School Diploma or Higher",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )

plot_wi
```

```{r, echo = FALSE, message = FALSE, warning= FALSE}

education <- get_acs(
  geography = "place",
  state = "NY",
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(education = estimate)


income <- get_acs(
  geography = "place",
  state = "NY",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(income = estimate)

city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(NAME = gsub(", New York", "", NAME)) %>%
  filter(!is.na(education) & !is.na(income))

```

```{r plot-new york, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8}
plot_ny <-ggplot(city_df, aes(x = education, y = income)) +
  geom_point(color = "blue", size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income Across New York Cities",
    x = "% of Adults with a High School Diploma or Higher",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )

plot_ny 
```



```{r, echo = FALSE, message = FALSE, warning= FALSE}

education <- get_acs(
  geography = "place",
  state = "WY",
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(education = estimate)


income <- get_acs(
  geography = "place",
  state = "WY",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(income = estimate)

city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(NAME = gsub(", Wyoming", "", NAME)) %>%
  filter(!is.na(education) & !is.na(income))

```

```{r plot-wyoming, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8}
plot_wy <-ggplot(city_df, aes(x = education, y = income)) +
  geom_point(color = "darkgreen", size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income Across Wyoming Cities",
    x = "% of Adults with a High School Diploma or Higher",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )

plot_wy
```






```{r, echo = FALSE, message = FALSE, warning= FALSE}

education <- get_acs(
  geography = "place",
  state = "CO",
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(education = estimate)


income <- get_acs(
  geography = "place",
  state = "CO",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(income = estimate)

city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(NAME = gsub(", Colorado", "", NAME)) %>%
  filter(!is.na(education) & !is.na(income))

```

```{r plot-colorado, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8}
plot_co <-ggplot(city_df, aes(x = education, y = income)) +
  geom_point(color = "darkorange", size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income Across Colorado Cities",
    x = "% of Adults with a High School Diploma or Higher",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )

plot_co
```




```{r, echo = FALSE, message = FALSE, warning= FALSE}

education <- get_acs(
  geography = "place",
  state = "LA",
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(education = estimate)


income <- get_acs(
  geography = "place",
  state = "LA",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(income = estimate)

city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(NAME = gsub(", Louisiana", "", NAME)) %>%
  filter(!is.na(education) & !is.na(income))

```

```{r plot-louisiana, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8}
plot_la <-ggplot(city_df, aes(x = education, y = income)) +
  geom_point(color = "cyan3", size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income Across Louisiana Cities",
    x = "% of Adults with a High School Diploma or Higher",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )
plot_la
```

```{r}

library(patchwork)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 19, fig.height = 12}
(plot_wi | plot_ny | plot_wy) /
(plot_co | plot_la)
```


## Analysis



```{r load-libraries, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(modelsummary)
```

```{r load-data, echo = FALSE, message = FALSE, warning = FALSE}

states <- c("WI", "NY", "WY", "CO", "LA")


education <- get_acs(
  geography = "place",
  state = states,
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(hs_graduation_rate = estimate)


income <- get_acs(
  geography = "place",
  state = states,
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(median_income = estimate)


city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(
    # Extract state name from the NAME column
    state = case_when(
      grepl("Wisconsin", NAME)   ~ "Wisconsin",
      grepl("New York", NAME)    ~ "New York",
      grepl("Wyoming", NAME)     ~ "Wyoming",
      grepl("Colorado", NAME)    ~ "Colorado",
      grepl("Louisiana", NAME)   ~ "Louisiana"
    ),
    
    NAME = gsub(", Wisconsin|, New York|, Wyoming|, Colorado|, Louisiana", "", NAME)
  ) %>%
  filter(!is.na(hs_graduation_rate) & !is.na(median_income) & !is.na(state))
```

```{r}

library(kableExtra)
```


```{r summary-table, echo = FALSE, message = FALSE, warning = FALSE}

city_df %>%
  group_by(state) %>%
  summarise(
    Cities = n(),
    `Avg HS Graduation Rate` = round(mean(hs_graduation_rate), 1),
    `Avg Median Income` = round(mean(median_income), 0)
  ) %>%
  kable(
    caption = "Summary of Data by State",
    col.names = c("State", "Number of Cities", "Average HS Graduation Rate (%)", "Average Median Household Income ($)"),
    format.args = list(big.mark = ",")
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "bordered"),
    full_width = TRUE
  )
```



```{r outcome-plot, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6}

ggplot(city_df, aes(x = median_income, fill = state)) +
  geom_histogram(bins = 40, alpha = 0.7, color = "white") +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Distribution of Median Household Income",
    subtitle = "Across all cities in Colorado, Louisiana, New York, Wisconsin, and Wyoming",
    x = "Median Household Income ($)",
    y = "Number of Cities",
    fill = "State",
    caption = "Source: U.S. Census Bureau, ACS 5-Year Estimates (2023)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )
```



```{r regression, echo = FALSE, message = FALSE, warning = FALSE}
# Run the regression
model <- lm(median_income ~ hs_graduation_rate, data = city_df)

# Display regression output nicely
modelsummary(
  model,
  stars = TRUE,
  coef_rename = c(
    "hs_graduation_rate" = "High School Graduation Rate (%)",
    "(Intercept)" = "Intercept"
  ),
  gof_map = c("nobs", "r.squared"),
  fmt = 0,
  title = "Regression: Effect of High School Graduation Rate on Median Household Income"
)
```