---
title: "PS270 Final Project"
author: "Amelia Alvarez"
description: "Income vs Education"
output:
  distill::distill_article:
    self_contained: false
---


## Project thoughts

I am interested in exploring data related to high school graduation rates in areas of poverty. I would like to look at low income areas and see if they have lower graduation rates compared to more affluent areas.

## Project Hypothesis

My research question examines how neighborhood income level is related to high school graduation rates. Are students in low-income areas less likely to graduate from high school compared to students in higher-income areas? My hypothesis is that high school students in low-income areas are less likely to graduate from high school than students in higher-income areas. This hypothesis is grounded in the thought that schools in low-income areas often receive fewer financial resources due to property tax-based funding structures. This can result in larger class sizes, fewer experienced teachers, limited access to advanced coursework, fewer counselors, and academic support services. From a behavioral perspective, economic hardship can also increase stress at home, require students to work part-time jobs, or reduce parental availability for academic support. These disadvantages systematically reduce academic performance and increase dropout risk. My explanatory variable of interest is the socioeconomic status of the area. This can be measured by median household income in the school district or ZIP code, the percentage of students eligible for free or reduced-price lunch, and the poverty rate in the census tract or district. These measures vary across different areas, which allows us to observe variation across schools or districts in income levels. The outcome variable of interest is high school graduation rates. This can be measured by the four-year cohort graduation rate, which is the percentage of students who graduate within four years of entering ninth grade, or by looking at the dropout rate. My hypothesis would be supported if the data show a negative relationship between area income and graduation rates and if schools in lower-income areas consistently show lower graduation rates. My hypothesis would be weakened if the data show no relationship between area income and graduation rates, or if, after controlling for other variables such as school size, urban or rural location, and racial composition, income is no longer statistically significant, which would challenge the source of my hypothesis.


```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(dplyr)
library(readr)
library(readxl)
library(ggplot2)
```
```{r, echo = FALSE, message = FALSE, warning = FALSE}
data1  <- readr::read_csv("data/kenosha.income.csv")
data2  <- readr::read_csv("data/kenosha_education(Sheet1).csv")
data3  <- readr::read_csv("data/madison.income.csv")
data4  <- readr::read_csv("data/madison_education(Sheet2).csv")
data5  <- readr::read_csv("data/westallis.income.csv")
data6  <- readr::read_csv("data/west allis_education(Sheet1).csv")
data7  <- read_excel("data/beloit.education.xlsx")
data8  <- read_excel("data/beloit.income.xlsx")
data9  <- read_excel("data/cedarburg.education.xlsx")
data10 <- read_excel("data/cedarburg.income.xlsx")
data11 <- read_excel("data/fitchburg.education.xlsx")
data12 <- read_excel("data/fitchburg.income.xlsx")
data13 <- read_excel("data/janesville.education.xlsx")
data14 <- read_excel("data/janesville.income.xlsx")
data15 <- read_excel("data/middleton.education.xlsx")
data16 <- read_excel("data/middleton.income.xlsx")
data17 <- read_excel("data/monroe.education.xlsx")
data18 <- read_excel("data/monroe.income.xlsx")
data19 <- read_excel("data/Oconomowoc.education.xlsx")
data20 <- read_excel("data/Oconomowoc.income.xlsx")
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Fix encoding on all datasets FIRST before anything else
all_datasets <- c("data1","data2","data3","data4","data5","data6","data7",
                  "data8","data9","data10","data11","data12","data13","data14",
                  "data15","data16","data17","data18","data19","data20")

for (name in all_datasets) {
  df <- get(name)
  df[] <- lapply(df, function(col) {
    if (is.character(col)) {
      col <- iconv(col, from = "", to = "UTF-8", sub = "byte")
      col <- iconv(col, from = "UTF-8", to = "UTF-8", sub = "")
    }
    col
  })
  assign(name, df)
}
```


```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Rename columns to be consistent across all datasets
data1  <- data1  %>% rename(label = Label, estimate = Estimate, margin_of_error = `Margin of Error`)
data2  <- data2  %>% rename(label = Label, margin_of_error = `margin of error`)
data3  <- data3  %>% rename(margin_of_error = `margin of error`)
data4  <- data4  %>% rename(label = Label, estimate = Estimate, margin_of_error = `Margin of error`)
data5  <- data5  %>% rename(label = lable, margin_of_error = `margin of error`)
data6  <- data6  %>% rename(label = Label, estimate = estimiate, margin_of_error = `margin of error`)
data7  <- data7  %>% rename(label = Label, margin_of_error = `margin of error`)
data8  <- data8  %>% rename(margin_of_error = `margin of error`)
data9  <- data9  %>% rename(label = Label, margin_of_error = `margin of error`)
data10 <- data10 %>% rename(margin_of_error = `margin of error`)
data11 <- data11 %>% rename(label = Label, margin_of_error = `margin of error`)
data12 <- data12 %>% rename(margin_of_error = `margin of error`)
data13 <- data13 %>% rename(label = Label, margin_of_error = `margin of error`)
data14 <- data14 %>% rename(margin_of_error = `margin of error`)
data15 <- data15 %>% rename(margin_of_error = `margin of error`)
data16 <- data16 %>% rename(margin_of_error = `margin of error`)
data17 <- data17 %>% rename(estimate = education, margin_of_error = `margin of error`)
data18 <- data18 %>% rename(margin_of_error = `margin of error`)
data19 <- data19 %>% rename(margin_of_error = `margin of error`)
data20 <- data20 %>% rename(margin_of_error = `margin of error`)
```



```{r, echo = FALSE, message = FALSE, warning = FALSE}
get_bachelors <- function(edu_data) {
  total_pop <- as.numeric(gsub("[,±]", "", edu_data$estimate[7]))
  bachelors <- as.numeric(gsub("[,±]", "", edu_data$estimate[16]))
  round((bachelors / total_pop) * 100, 1)
}

get_high_income <- function(inc_data) {
  inc_data %>%
    filter(grepl("\\$75,000 to \\$99,999", label, ignore.case = TRUE)) %>%
    mutate(estimate = as.numeric(gsub("[%,\u00b1]", "", estimate))) %>%
    pull(estimate) %>%
    first()
}

# Build city dataframe
city_df <- data.frame(
  city = c("Kenosha", "West Allis", "Beloit", "Cedarburg",
           "Fitchburg", "Janesville", "Middleton", "Monroe", "Oconomowoc"),

  education = c(
    get_bachelors(data2),
    get_bachelors(data6),
    get_bachelors(data7),
    get_bachelors(data9),
    get_bachelors(data11),
    get_bachelors(data13),
    get_bachelors(data15),
    get_bachelors(data17),
    get_bachelors(data19)
  ),

  income = c(
    get_high_income(data1),
    get_high_income(data5),
    get_high_income(data8),
    get_high_income(data10),
    get_high_income(data12),
    get_high_income(data14),
    get_high_income(data16),
    get_high_income(data18),
    get_high_income(data20)
  )
)

```





```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 7}
ggplot(city_df, aes(x = education, y = income, label = city)) +
  geom_point(color = "darkgreen", size = 4) +
  geom_text(vjust = -0.8, size = 3.5, fontface = "bold") +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income in Wisconsin Cities",
    x = "% of Adults with a Bachelor's Degree or Higher",
    y = "% of Households Earning $75,000 to $99,999"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )
```

## Analysis

At first glance, the trend line suggests a slight negative relationship between the two variables, meaning cities with higher education rates do not necessarily have higher groups of households in the $75,000–$99,999 income bracket. This is likely reflecting the fact that highly educated cities like Middleton and Oconomowoc have more households earning above $99,999, pushing residents out of this specific income bracket rather than into it.  Wealthier, more educated cities may have fewer households in the $75,000–$99,999 range because their incomes skew even higher.

Cities like Monroe, Beloit, and West Allis cluster on the left side of the chart with lower education rates but mid-range income concentrations, while Middleton and Oconomowoc sit on the far right with high education rates but lower concentrations in this specific income band. Kenosha stands out as having both a low education rate and a low income concentration.


```{r}
install.packages("tidycensus")
```

```{r}
library(tidyverse)
library(tidycensus)
```



```{r}
census_api_key("1956434ce467efbcc6650c2a36eb9a5c4f758f22", install = TRUE)
```
```{r}
education <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP02_0068P",
  year = 2023,
  survey = "acs5"
)
```


```{r}
income <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
)
```


```{r, echo = FALSE, message = FALSE, warning= FLASE}

library(tidycensus)
library(tidyverse)
library(ggplot2)

census_api_key("1956434ce467efbcc6650c2a36eb9a5c4f758f22")

```


```{r, echo = FALSE, message = FALSE, warning= FLASE}

education <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP02_0066P",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(education = estimate)


income <- get_acs(
  geography = "place",
  state = "WI",
  variables = "DP03_0062",
  year = 2023,
  survey = "acs5"
) %>%
  select(GEOID, NAME, estimate) %>%
  rename(income = estimate)

city_df <- inner_join(education, income, by = c("GEOID", "NAME")) %>%
  mutate(NAME = gsub(", Wisconsin", "", NAME)) %>%
  filter(!is.na(education) & !is.na(income))

```


```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8}
ggplot(city_df, aes(x = education, y = income)) +
  geom_point(color = "purple", size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red",
              linetype = "dashed", alpha = 0.15) +
  labs(
    title = "Education vs. Income Across Wisconsin Cities",
    x = "% of Adults with a High School Diploma or Higher",
    y = "Median Household Income ($)",
    caption = "Source: U.S. Census Bureau, American Community Survey 5-Year Estimates (2023)"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray50"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 40, 20, 40)
  )
```

